EDIpack2: interoperable Lanczos-based solver for generic quantum impurity problems

### 2025-04-10 02_install.tex
Updated draft. EDIpy2 to be completed. Next skip to examples.

\subsection{Structure}\label{SSecStruct}
\NAME is a modular library which contains three principal structures.
At the core is the exact diagonalization solver: \NAME.
Next there is a {\tt EDIpack2ineq} which extends application to the case of multiple
inequivalent impurity problems. Finally, there we provide a Fortran-C/C++
interface, which enables development of additional API or
interoperability with other, external, software.

@@

\subsection{Dependencies}
\NAME essentially depends on two external libraries.
\begin{itemize}
\item {\bf SciFortran}: an open-source Fortran library to support
  mathematical and scientific software development. 
\item {\bf MPI} (optional): a distributed memory parallel communication layer with support to modern Fortran compiler.
\end{itemize}

@@

\subsection{Build and Install}
\subsubsection{Source}
The software can be installed from source as follows. The source can
be retrieved directly from its GitHub repository, for instance using:
\begin{lstlisting}[style=mybash]
git clone https://github.com/edipack/EDIpack2.0 EDIpack2
\end{lstlisting}
Then, assuming to be in the software directoru, a conventional
out-of-source building is performed using two different compilations
backends.

@@

\subsection{OS Loading}
In oder to avoid possible conflicts or require administrative
privileges, the building step results get installed by default in a user home
directory, specified by the CMake variable {\tt PREFIX}.
In doing so, however, one misses the chance of automatic loading into
the operative system.

>>>>

\subsection{Structure}\label{sSecInstallStructure}
\NAME is a modular library which contains three principal structures.
At the core is the exact diagonalization solver: \NAME.
Next there is a {\tt EDIpack2ineq} which extends application to the case of multiple
inequivalent impurity problems. Finally, there we provide a Fortran-C/C++
interface, which enables development of additional API or
interoperability with other, external, software.

@@

\subsection{Dependencies}\label{sSecInstallDependencies}
\NAME essentially depends on two external libraries.
\begin{itemize}
\item {\bf SciFortran}: an open-source Fortran library to support
  mathematical and scientific software development. 
\item {\bf MPI} (optional): a distributed memory parallel communication layer with support to modern Fortran compiler.
\end{itemize}

@@

\subsection{Build and Install}\label{sSecInstallBuildInstall}
\subsubsection{Source}
The software can be installed from source as follows. The source can
be retrieved directly from its GitHub repository, for instance using:
\begin{lstlisting}[style=mybash]
git clone https://github.com/edipack/EDIpack2.0 EDIpack2
\end{lstlisting}
Then, assuming to be in the software directoru, a conventional
out-of-source building is performed using two different compilations
backends.

@@

\subsection{OS Loading}\label{sSecInstallOSloading}
In oder to avoid possible conflicts or require administrative
privileges, the building step results get installed by default in a user home
directory, specified by the CMake variable {\tt PREFIX}.
In doing so, however, one misses the chance of automatic loading into
the operative system.

### 2025-04-10 04_cbinding.tex
Updated draft. EDIpy2 to be completed. Next skip to examples.

\section{Interoperability}
The recent growing availability of state-of-the-art software dedicated
to the solution of quantum impurity problems using different methods
poses a serious challenge to test accuracy and realiability of the
results.
As such, complex softwares are requested to develop a higher level of
interoperability, i.e. the possibility to operate with other software
possibly using different programming language.
Modern Fortran, which is the language of choice for \NAME, since many
years supports the standardized generation of procedures and
variables that are interoperable with C.

\subsection{C-bindings}\label{sSecCbind}
The interoperability with C language is encoded in the
{\tt ISO\_C\_BINDING} module part of Fortran
standard since 2003. The module contains definitions of named
constants, types and procedures for C interoperability.
Alongside, the second key feature essential to expose to C any Fortran
entity is the {\tt BIND(C)} intrinsic function.

\subsection{EDIpy, the Python API}\label{SecPyAPI}

\subsubsection{Build from source}
EDIpy2 is available as a stand-alone
module which depends on both \NAME and SciFortran. The package can be
obtained from the repository
\href{https://github.com/EDIpack/EDIpy2.0}{EDIpy2}.

@@

\subsubsection{Anaconda}
As for EDIpack2, also the Python API in EDIpy2 are available through
Anaconda packaging. In this case the resolution of the dependencies is
taken care from Conda itself:

\begin{lstlisting}[style=mybash]
conda create -n edipack
conda activate edipack
conda install -c conda-forge -c edipack edipack2
\end{lstlisting}

\subsection{TRIQS interface}\label{SecEDI2Triqs}
A purely Python \NAME to Triqs interface is available, leveraging on
the C-bindings and Python API. The corresponding module depends on
\NAME (which ultimately depends on SciFortran) and Triqs.
Assuming the two software are correctly installed in the OS, the
EDIpack2Triqs interface is installed as follows:

@@

\subsection{W2Dynamics interface}\label{SecEDI2W2Dyn}

\subsection{EDIjl, the Julia API}\label{SecJlAPI}

>>>>

\section{Interoperability}\label{SecInterop}
The recent growing availability of state-of-the-art software dedicated
to the solution of quantum impurity problems using different methods
poses a serious challenge to test accuracy and realiability of the
results.
As such, complex softwares are requested to develop a higher level of
interoperability, i.e. the possibility to operate with other software
possibly using different programming language.
Modern Fortran, which is the language of choice for \NAME, since many
years supports the standardized generation of procedures and
variables that are interoperable with C.

\subsection{C-bindings}\label{sSecInteropCbindings}
The interoperability with C language is provided by the
{\tt ISO\_C\_BINDING} module, which is part of the Fortran
standard since 2003. The module contains definitions of named
constants, types and procedures for C interoperability.
Alongside, a second key feature essential to expose to C any Fortran
entity is the {\tt BIND(C)} intrinsic function.
In \NAME we exploit these features of the language to provide a
complete interface from the Fortran code to C/C++, i.e. {\tt
  EDIpack2C}.

\subsubsection{Installation}\label{sSecInteropCbindingsInstallation}
The C-binding module in included in the build process of \NAME and
compiled into a dynamical library {\tt
  libedipack2\_cbinding.so/.dylib}. As dicussed in the
\secu{sSecInstallBuildInstall}, the support for inequivalent
impurities is configured at the build level and propagates to the
C-bindings library as well. An exported variable {\tt has\_ineq} is
defined and export to C/C++ as a way to query the presence of
inequivalent impurities support. 
The generated library and header files get installed in the include
directory at the prefix location, specified during configuration
step. The corresponding path is added to the environment variable {\tt
  LD\_LIBRARY\_PATH}, valid of any Unix/Linux system, via any of the
loading methods outlines in \secu{sSecInstallOSloading}.

\subsubsection{Implementation}\label{sSecInteropCbindingsImplementation}
The interface layer is contained in the Fortran module {\tt
  EDIpack2\_C}. This contains a common part and two set of functions,
one to interface the procedures from \NAME and a second
one to extend the interface to the inequivalent impurities case.
Specifically, the implemented interface functions expose to C through
{\tt bind(C)} statement a number of
procedures composing the Fortran API of \NAME, i.e. contained in {\tt
  ED\_MAIN}. The procedures and shared variables can be divided in four main groups:

\paragraph{{\bf Variables}}
A number of relevant input and shared variables which are normally
required to setup or to control the calculation are interfaced to C
directly in the \NAME modules {\tt ED\_INPUT\_VARIABLES} using {\tt
  bind(C)} constructs. These are implicitly loaded into the C-binding module {\tt
  EDIpack2\_C} through the Fortran {\tt USE EDIPACK2} statement and
then further interfaced in the C++ header file.

\paragraph{{\bf Main}} This group contains interface to the exact
digonalization solver, interfacing the solver itself {\tt
  solve\_site/ineq} as well as its initialization
{\tt init\_solver\_site/ineq} and finalization {\tt
  finalize\_solver} procedures. It also includes the functions used to
set the non-interacting part of the impurity Hamiltonian
$h^0_{\a\b\s\s'}$ through the functions {\tt set\_Hloc}.

\paragraph{{\bf Bath}} In this group we implement a number of
procedures dealing with bath initialization, symmetry operations and
optimization. In particular it contains the function returning the
dimension of the bath array on the user side {\tt
  get\_bath\_dimension} as well as the setup of the matrix basis
$\vec{\Gamma}$ for the replica/general bath via different instances of
{\tt set\_Hreplica/general\_{site,lattice}\_d{3,5}}. 
The crucial part in a DMFT framework is the optimization of the bath
through conjugate-gradient algorithm. To this end we interfaced a
number of functions {\tt
  chi2\_fitgf\_{single,lattice}\_{normal,superc}\_n{3,4,5,6}} which
cover all the cases supported in \NAME. Note that, because the actual
optimization is still performed through the Fortran code, no changes
apply to the outcome of this step.

\paragraph{{\bf Input/Output}}
The input and output part of the software is interfaced in this group
of functions. In particular, {\tt read\_input} expose to C the input
reading procedure of \NAME which set essentially set all the internal
variables of \NAME.
Next, in {\tt edipack{2ineq}\_c\_binding\_io} we interface all the
functions implementing the communication from the \NAME instance to
the user, namely those to retrieve static observables (e.g. {\tt
  get\_dens}), the impurity Green's functions and self-energies
(e.g. {\tt get\_Sigma\_{site,lattice}\_n{3,5}}), the
impurity susceptibilities (e.g. {\tt get\_spinChi}), the impurity reduced density matrix as well
as the non-interacting Green's and hybridization functions starting from the
user bath array.

Finally, in order to ensure compatibility to C and C++ language the
entire set of C-binding functions and variables is enclosed into a
header file {\tt edipack2\_cbinding.h}.

% ##################################################################
% ##################################################################

\subsection{EDIpy2, the Python API}\label{sSecInteropEDIpy}
As a first application of the \NAME C-bindings we implemented a
complete Python interface, i.e. EDIpy2. This is a Python module which
enables access to all the library features and unlock implementation of
further interfaces of \NAME as a plug-in solver for external Python
based software.

\subsubsection{Installation}\label{sSecInteropEDIpyInstallation}
EDIpy2 is available as a stand-alone
module which only depends on \NAME and SciFortran.
The Python package can be obtained from the repository
\href{https://github.com/EDIpack/EDIpy2.0}{EDIpy2}.

@@

\paragraph{PyPi}
The EDIpy2 package is also available in PyPi at
\href{https://pypi.org/project/edipy2/}{pypi.org/edipy2}. As such
EDIpy2 can be conveniently installed in any system supporting {\tt pip} as:

\begin{lstlisting}[style=mybash]
pip install edipy2
\end{lstlisting}

\paragraph{Anaconda}
As for \NAME, the Python API in EDIpy2 is available through
Anaconda packaging for Unix/Linux systems. Packages are available for
Python version $>3.10$. The EDIpack2.0 package contains the {\tt
  edipy2} Python module as well as the EDIpack2.0 and SciFortran
libraries. In this case the resolution of the dependencies is
taken care from Conda itself. 
Using conda the installation reads:
\begin{lstlisting}[style=mybash]
conda create -n edipack
conda activate edipack
conda install -c conda-forge -c edipack edipack2
\end{lstlisting}

Once installed, any loading of the {\tt edipy2} module resolves in the
attempt to load the dynamic library {\tt
  libedipack2\_cbinding.so/.dylib} containing the Fortran-C bindings
for \NAME. By default the library is searched proceeds as follow: 
\begin{enumerate}
\item First the user can override the location of the library
  (determined during \NAME build configuration) by exporting an
  environment variable called {\tt EDIPACK\_PATH} .
\item By default, the Python module detects the position of the
  Fortran libraries via {\tt pkg-config}. Any of the OS loading method
  outlined in \secu{sSecInstallOSloading} automatically push the
  correct configuration to the {\tt PKG\_CONFIG\_PATH}. 
\item As a last resort the environment variables {\tt
    LD\_LIBRARY\_PATH} and {\tt DYLD\_LIBRARY\_PATH} are analyzed to
  retrieve the correct position. 
\end{enumerate}
If none of the previous attempts succeeds, the module will not not load correctly and an error message will be printed.

\subsubsection{Implementation}\label{sSecInteropEDIpyImplementation}
The Python API provided in the edipy2 module consists essentially of a
suitable class called for convenience {\tt global\_env}.
The class incorporates all the global variables inherited from \NAME
C-bindings library and implements a number of interface functions
leveraging the Python duck typing to \NAME.  
The variables and the functions of \NAME are exposed to the user and
are accessed as properties and
methods of the {\tt global\_env} claass, which needs to be imported at the beginning
of the python script, along with other useful modules. For instance Numpy is
necessary, while mpi4py is strongly recommended.

\begin{lstlisting}[language=python,  frame=lines]
import numpy as np
import mpi4py
from mpi4py import MPI
from edipy2 import global_env as ed
import os,sys
\end{lstlisting}

The EDIpy2 supports the solution of problems with independent
impurities, interfacing with the the EDIpack2ineq extension of the
library, if present. Should the inequivalent impurities package not be
built, the Python module silently disable the support to it, so that
invoking any related procedure  will result in a {\tt RuntimeError}.
The user can check the availability of the inequivalent impurities
interface by querying the value of  {\tt edipy2.global\_env.has\_ineq}.

The implementation of the Python API includes two main parts, namely
the global variables and a number of functions.

\paragraph{{\bf Global variables}}

\paragraph{{\bf Functions}}

% ##################################################################
% ##################################################################

\subsection{TRIQS interface}\label{sSecInteropTRIQS}
{\color{red} Igor }
A purely Python \NAME to Triqs interface is available, leveraging on
the C-bindings and Python API. The corresponding module depends on
\NAME (which ultimately depends on SciFortran) and Triqs.
Assuming the two software are correctly installed in the OS, the
EDIpack2Triqs interface is installed as follows:

@@

\subsection{W2Dynamics interface}\label{sSecInteropW2DYN}
{\color{red} Giorgio, Alexander,?}

\subsection{EDIjl, the Julia API}\label{sSecInteropEDIjl}
{\color{red} Lorenzo?}
